Investigating magnetic dynamo evolution with \TESS\ field dwarfs

### 2017-10-06 ms.tex
some more editing and shortening

Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star's rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star's surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\% or less) quasi-periodic changes in the stellar
brightness.
Thousands of photometric rotation periods have been measured from \Kepler\ and
\ktwo\ light curves, but these surveys focused on small areas of the sky.
\tess\ will open unexplored parts of the galaxy to rotational evolution
studies.
Some outstanding mysteries regarding the nature of magnetic braking were
revealed, but thus-far unanswered, by \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars
\citep{van-saders2016}.
To answer these questions {\bf we propose to provide a catalog of $\sim$ 1
million rotation periods for \TESS\ Full Frame Image (FFI) stars.}

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport found this period bimodality extends throughout all
masses in the \Kepler\ rotation sample for nearby stars \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The \TESS\ FFI targets will provide the ideal dataset to test these two
scenarios explaining the appearance of a period bimodality.
If the bimodal period distribution reflects a discontinuous age distribution,
the feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the \TESS\ FFI targets to the upcoming second data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactocentric position.
For the all-southern-hemisphere  analysis we will focus on F and G stars since
the 27.4 day baseline of \TESS\ fields will limit measurable rotation periods
to less than around 15 days.
The longer time coverage in the Continuous Viewing Zone (CVZ) will allow us to
study the rotation period bimodality in K and M dwarfs, since the bimodality
appears at longer rotation periods for these lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
coverage necessary to map the period bimodality across the galaxy.}

@@

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it does
not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new rotation periods, with the maximum
measurable period strongly depending on the length of time it is observed for.
Of particular interest are the potentially long rotation periods measurable
for stars that fall in the CVZ.
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of \TESS\ field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many \tess\ FFI targets will have proper motions, parallaxes, positions and
radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action-angle coordinates of the stars, both of which are age
indicators.
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets in the CVZ.}

\citet{oh2016} identified over 10,000 comoving pairs of stars using proper
motions from the first \Gaia\ data release.
Of these, 18 dwarfs and subgiants in 9 pairs have \RAVE\ radial velocities
consistent with being comoving and are already in the Candidate Target List.
Following \citet{campante2017}, these stars have oscillation detection
probabilities greater than 50\%, based on their \RAVE-on T$_{\mathrm{eff}}$,
$\log g$s and radii \citep{casey2017}.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list with this proposal.
The stars in these pairs are likely recently disrupted binaries so we
expect them to have similar ages and measuring their rotation periods will
therefore provide a test of gyrochronology.
An additional 5000 bright comoving pair candidates lie in the southern
ecliptic hemisphere and will be ideal FFI targets for gyrochronology studies.
{\bf We will use the rotation periods of comoving stars to quantify the
accuracy and precision of gyrochronology.}

@@

Based on \citet{sullivan2015}, we expect around 10 million \TESS\ FFI
targets brighter than 15th magnitude in the southern ecliptic hemisphere.
We intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic, Gaussian process
period analysis on the high priority targets.
High priority targets will include: comoving pairs, stars in the CVZ
(especially those with spectra), stars in TGAS, stars at large distances, plus
any potential asteroseismic dwarfs and exoplanet hosts.
The number of high priority targets with probabilistic rotation periods will
be limited by computation power, but we conservatively estimate that we can
produce full rotation period posterior PDFs for $\sim$ 30,000 stars.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 15th
magnitude.
We note, however that not all stars will have measurable rotation periods,
even if they are bright enough not all stars show rotational variability in
their light curves either because they are pole-on or inactive.
We will vet incorrect rotation period measurements using a combination of a
Lomb-Scargle periodogram and an autocorrelation function by
training on a set of light curves with a range of variability amplitudes and
periods.
Based on the fraction of stars with measurable rotation periods in
\citet{mcquillan2014}, and the more restrictive baseline of \tess\ relative to
\kepler, we predict that the final catalog will contain around one million
rotation periods.

@@

\ssfigure{Kalesalad.png}{0.5}{%
All stars observed during K2 campaign 4, plotted according to their
equatorial coordinates and colored by their preliminary rotation period.
These rotation periods were measured using a simple ACF method, applied to
    everest (Luger et al., 2015) light curves.
\label{fig:kalesalad}
}

\paragraph{Expected Impact}
We will provide rotation period posterior probability density functions and
best fit estimates for $\sim$ 30,000 high priority FFI targets rotation
periods and point estimates for $\sim$ 10 million lower priority FFI stars
using a combination of Lomb-Scargle periodogram and autocorrelation
techniques.
We will release an open source python package for measuring rotation periods
of \tess\ FFI targets plus code for inferring stellar ages from rotation
periods, based on a new gyrochronology relation calibrated using \TESS --
\HERMES spectra, \Gaia\ kinematics and comoving pairs.

\paragraph{Budget Justification}
PI Angus will employ two undergraduate students for 3-4 months who will assist
in extracting light curves from the \tess\ FFIs, measuring rotation periods
and building a rotation period catalog.
These students will also be involved in the scientific project of their
choosing: either the rotation period bimodality, gyrochronology of FGKM stars,
or gyrochronology of comoving pairs.
The remaining budget will be used for a bringing all Co-Is together in order
to make and execute analysis plans and for presenting the results at
conferences.
\paragraph{References}

>>>>

Of the measurable properties for a large ensemble of field stars, rotation
periods contain the most information about stellar age and provide the best
leverage for advancing our knowledge of galactic archeology as well as
exoplanet population demographics via gyrochronology.
Angular momentum is carried away though magnetically driven stellar winds,
which slows the star's rotation over cosmic time.
This rotation-based `clock' is known as gyrochronology.
Cool spots on the star's surface rotate in-to and out-of view, creating small
amplitude ($\pm$ 1\% or less) quasi-periodic changes in the stellar
brightness from which a rotation period can be measured.
Thousands of photometric rotation periods have been measured from \Kepler\ and
\ktwo\ light curves this way, but these surveys focused on small areas of the
sky whereas \tess\ will open unexplored parts of the galaxy to rotational
evolution studies.
Some outstanding mysteries regarding the nature of magnetic braking were
revealed, but thus-far unanswered, by \kepler.
Firstly, a mysterious gap in the rotation period distribution of \Kepler\
dwarfs requires either a sharp transition in magnetic dynamo geometry or a gap
in the local star formation history as an explanation \citep{mcquillan2014,
davenport2017}.
Secondly, a transitioning magnetic dynamo appears to be responsible for
inefficient magnetic braking at old ages in Solar-mass stars
\citep{van-saders2016}.
To shed light on this behavior {\bf we propose to provide a catalog of $\sim$
1 million rotation periods for \TESS\ Full Frame Image (FFI) stars, to map the
rotation period bimodality as a function of galactic position and to calibrate
gyrochronology using \tess--\hermes\ spectra, \gaia\ kinematics and \gaia\
comoving pairs.}

@@

The bimodal period distribution among field stars discovered by
\citet{mcquillan2013}, is shown in Figure \ref{fig:davenport}.
Recently, Co-I Davenport found this period bimodality extends throughout all
masses in the \Kepler\ rotation sample for nearby stars \citep{davenport2017}.
This feature could either reflect a drop in the star formation rate around 600
Myr ago or could be explained by a previously unknown variation in the
spin-down evolution for low-mass stars.
The \TESS\ FFI targets will provide the ideal dataset to test these two
scenarios.
If the period bimodality reflects a discontinuous age distribution, the
feature should be local and may disappear at greater distances or along
different lines of sight.
However, if the bimodality is truly due to a transition point in the spin-down
evolution at young ages, there should be little to no variation in the feature
with galactic position.
We will match the \TESS\ FFI targets to the upcoming second data
release from the Gaia mission \citep{perryman2001} to map the rotation period
distribution as a function of galactic position.
Outside the Continuous Viewing Zone (CVZ) we will focus on F and G stars since
the 27.4 day baseline of \TESS\ fields will limit measurable rotation periods
to less than around 15 days.
The longer time coverage in the CVZ will allow us to study the rotation period
bimodality in K and M dwarfs too, since the bimodality appears at longer
rotation periods for these lower mass stars.
Unlike \kepler\ and \ktwo\ light curves, {\bf \TESS\ provides the all-sky
coverage necessary to map the period bimodality across the galaxy.}

@@

Although the classical spin-down law of \citet{skumanich1972} (Period
$\propto$ Age$^{1/2}$) holds for all open clusters with measured periods, it
does not appear to describe old field stars \citep{angus2015, van-saders2016}.
\citet{van-saders2016} found that including a transition to a weakened magnetic
braking regime in the gyrochronology models, at a critical Rossby number,
provides an improved fit to the data.
Further calibration is needed however; a lack of rotation periods and reliable
ages for old and low-mass stars leaves the rotational evolution of stars older
than the Sun relatively unexplored.
\TESS\ will provide thousands of new {\it long } rotation periods for stars
that fall in the CVZ.
Many of these will have spectra from the \TESS --\HERMES\ survey, from which
it will be possible to calculate isochronal ages, enabling further calibration
of the gyrochronology relations.
{\bf We propose to calibrate gyrochronology at old ages for FGKM stars using
the isochronal ages of \TESS--\hermes\ stars.}
We will also use the rotation periods of \TESS\ field dwarfs over the whole sky
to test the gyrochronology relations using galactic kinematics.
Many \tess\ FFI targets will have proper motions, parallaxes, positions and
radial velocities published in the second \Gaia\ data release.
These parameters provide the information necessary to calculate galactocentric
positions and action-angle coordinates of the stars, both of which are age
indicators.
{\bf We propose to calibrate the gyrochronology relations at old ages for FGKM
stars using vertical actions of \Gaia\ DR2 targets.}
Since one of the few ways to accurately age-date fully convective, late M
dwarfs is via kinematics, these new relations will help to infer ages for all
stars with M$<$ 0.35M$_\odot$, to which gyrochronology cannot be applied.

\citet{oh2016} identified over 10,000 comoving pairs of stars using proper
motions from the first \Gaia\ data release.
Of these, 18 dwarfs and subgiants in 9 pairs have \RAVE\ radial velocities
consistent with being comoving and are already in the Candidate Target List.
Following \citet{campante2016} we calculate that these stars have oscillation
detection probabilities greater than 50\%, based on their \RAVE-on
T$_{\mathrm{eff}}$, $\log g$ and radii \citep{casey2017}.
Since these stars would make interesting targets for both asteroseismology and
exoplanet search, we include a list with this proposal.
The stars in these pairs are likely recently disrupted binaries so we
expect them to have similar ages and measuring their rotation periods will
therefore provide a test of gyrochronology.
An additional 5000 bright comoving pair candidates lie in the southern
ecliptic hemisphere and will be ideal FFI targets for gyrochronology studies.
{\bf We will use the rotation periods of comoving stars to quantify the
accuracy and precision of gyrochronology.}

@@

Based on \citet{sullivan2015}, we expect around 10 million \TESS\ FFI
targets brighter than 15th magnitude in the southern ecliptic hemisphere.
We intend to prioritize certain groups of stars for period analysis,
calculating a simple (and computationally fast) Lomb-Scargle periodogram for
the low priority objects and running a full probabilistic, Gaussian process
period analysis on the high priority targets.
High priority targets will include: comoving pairs, stars in the CVZ
(especially those with spectra), stars in TGAS, stars at large distances, plus
potential asteroseismic dwarfs and exoplanet hosts.
The number of high priority targets with probabilistic rotation periods will
be limited by computation power, but we conservatively estimate that we can
produce full rotation period posterior PDFs for $\sim$ 30,000 stars.
Based on the expected photometric precision of \tess\ \citep{sullivan2015}, we
expect to be able to measure rotation periods for FFI stars down to 15th
magnitude.
We note however that not all stars will have measurable rotation periods:
even if they are bright enough not all stars show rotational variability in
their light curves either because they are pole-on or inactive.
We will vet incorrect rotation period measurements using a combination of a
Lomb-Scargle periodogram and an autocorrelation function by
training on a set of light curves with a range of variability amplitudes and
known periods.
Based on the fraction of stars with measurable rotation periods in
\citet{mcquillan2014}, and the more restrictive baseline of \tess\ relative to
\kepler, we predict that the final catalog will contain around one million
rotation periods.

@@

\ssfigure{Kalesalad.png}{0.3}{%
All stars observed during K2 campaign 4, plotted according to their
equatorial coordinates and colored by their preliminary rotation period.
These rotation periods were measured using a simple ACF method, applied to
    everest (Luger et al., 2015) light curves.
\label{fig:kalesalad}
}

\paragraph{Expected Impact}
We will provide rotation period posterior probability density functions and
best fit estimates for $\sim$ 30,000 high priority FFI targets rotation
periods and point estimates for $\sim$ 1 million lower priority FFI stars
using a combination of Lomb-Scargle periodogram and autocorrelation
techniques.
We will release an open source python package for measuring rotation periods
of \tess\ FFI targets plus code for inferring stellar ages from rotation
periods, based on a new gyrochronology relation calibrated using \TESS--
\HERMES\ spectra, \Gaia\ kinematics and comoving pairs.

\paragraph{Budget Justification}
PI Angus will employ two undergraduate students for 3-4 months who will assist
in extracting light curves from the \tess\ FFIs and building a rotation period
catalog.
These students will also be involved in the scientific project of their
choosing: either the rotation period bimodality, gyrochronology of FGKM stars,
or gyrochronology of comoving pairs.
The remaining budget will be used for bringing all Co-Is together in order
to make and execute analysis plans and for presenting the results at
conferences.
\paragraph{References}
